---
title: "Social Media Usage and Well-being"
author: "NISHA MANDAL"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This report analyzes the relationship between social media usage and
lifestyle/well-being indicators such as screen time, sleep, loneliness,
and job satisfaction, using a survey dataset of working adults.

---

## 1. Data Importing

```{r}
# Read the CSV file into R (must be in the same folder as this .Rmd file)
social_data <- read.csv("mental_health.csv")

# Preview the first few rows
head(social_data)
```

**Explanation:** `read.csv()` loads the dataset into a data frame called
`social_data`. `head()` shows the first 6 rows so we can confirm it loaded
correctly.

---

## 2. Data Structure and Overview

```{r}
str(social_data)
```

```{r}
summary(social_data[, c("Age", "Screen_Time_Hours_Day", "Social_Media_Hours_Day",
                         "Sleep_Hours_Night", "Loneliness", "Job_Satisfaction")])
```

**Explanation:** `str()` shows the data type of every column in the full
dataset. `summary()` is applied only to the columns relevant to our topic
(age, screen time, social media hours, sleep, loneliness, and job
satisfaction) to keep the focus on social media and well-being.

---

## 3. Data Preprocessing and Cleaning

```{r}
# Check for missing values in the columns we care about
colSums(is.na(social_data[, c("Age", "Gender", "Screen_Time_Hours_Day",
                               "Social_Media_Hours_Day", "Sleep_Hours_Night",
                               "Loneliness", "Job_Satisfaction")]))
```

```{r}
# Check for duplicate rows
sum(duplicated(social_data))
```

**Explanation:** `colSums(is.na(...))` counts missing values in our
relevant columns. We also check for duplicate rows to ensure data quality.
This dataset does not require date conversion since it is a one-time survey
(no time-series column).

---

## 4. Data Analysis and Statistical Exploration

```{r}
# Gender distribution among respondents
table(social_data$Gender)
```

```{r}
# Central tendency and spread of daily social media usage (hours)
median(social_data$Social_Media_Hours_Day)
var(social_data$Social_Media_Hours_Day)
sd(social_data$Social_Media_Hours_Day)
```

```{r}
# Average social media usage by gender
usage_by_gender <- aggregate(Social_Media_Hours_Day ~ Gender, data = social_data, mean)
usage_by_gender
```

```{r}
# Correlation between social media usage and loneliness
cor(social_data$Social_Media_Hours_Day, social_data$Loneliness)
```

```{r}
# Correlation between social media usage and sleep hours
cor(social_data$Social_Media_Hours_Day, social_data$Sleep_Hours_Night)
```

```{r}
# Correlation between screen time and job satisfaction
cor(social_data$Screen_Time_Hours_Day, social_data$Job_Satisfaction)
```

**Explanation:** `table()` shows how many respondents fall into each
gender category. `median()`, `var()`, and `sd()` describe how spread out
daily social media usage is across respondents. `aggregate()` compares
average usage between genders. `cor()` checks whether higher social media
usage relates to loneliness or reduced sleep, and whether general screen
time relates to job satisfaction.

---

## 5. Graphs and Plots

```{r}
# Pie chart: gender distribution
gender_counts <- table(social_data$Gender)

pie(
  gender_counts,
  main = "Gender Distribution of Respondents",
  labels = paste(names(gender_counts), gender_counts)
)
```

```{r}
# Bar chart: average social media usage by gender
barplot(
  usage_by_gender$Social_Media_Hours_Day,
  names.arg = usage_by_gender$Gender,
  main = "Average Daily Social Media Usage by Gender",
  xlab = "Gender",
  ylab = "Hours per Day",
  col = "steelblue"
)
```

```{r}
# Scatter plot: Social media usage vs Loneliness
plot(
  social_data$Social_Media_Hours_Day,
  social_data$Loneliness,
  main = "Social Media Usage vs Loneliness",
  xlab = "Social Media Hours per Day",
  ylab = "Loneliness Score",
  pch = 19,
  col = "darkgreen"
)
```

```{r}
# Scatter plot: Social media usage vs Sleep hours
plot(
  social_data$Social_Media_Hours_Day,
  social_data$Sleep_Hours_Night,
  main = "Social Media Usage vs Sleep Hours",
  xlab = "Social Media Hours per Day",
  ylab = "Sleep Hours per Night",
  pch = 19,
  col = "darkorange"
)
```

**Explanation:** The pie chart shows the gender split of respondents. The
bar chart compares average social media usage between genders. The two
scatter plots visualize whether more social media use relates to higher
loneliness or reduced sleep.

---

## 6. Findings and Conclusion

*(Fill this in after running the chunks above and looking at your actual
numbers/plots — a few example points to build on:)*

- The average respondent spends **[X]** hours per day on social media,
  with a standard deviation of **[X]**, showing [wide/narrow] variation
  across individuals.
- **[Gender]** respondents reported [higher/lower] average social media
  usage than **[other gender]**.
- The correlation between social media usage and loneliness was **[value]**,
  indicating a [strong/moderate/weak] relationship — meaning more time on
  social media [does/does not] clearly relate to feeling lonelier.
- The correlation between social media usage and sleep hours was **[value]**,
  suggesting that higher usage [does/does not] noticeably reduce sleep.
- The correlation between screen time and job satisfaction was **[value]**,
  showing a [strong/moderate/weak] relationship between overall screen time
  and how satisfied people feel at work.

**Conclusion:** This analysis explored how social media usage relates to
well-being indicators like loneliness and sleep. While some relationships
were observed, correlation alone cannot establish that social media use
directly causes these outcomes — other lifestyle and personal factors
likely also play a role. Further research with controlled variables would
be needed to draw stronger conclusions.
